---
title: "DAC 2022"
author: "TA session - Arthur Cheib"
date: "2/17/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Where does this lecture/code comes from?
This amazing code was provided by Thomas Lin Pedersen. He prepared this code for his famous workshop on ggplot2 (https://www.youtube.com/watch?v=h29g21z0a68&t=0s)

```{r}
### Schedule for today

## 1) Go through some of the "layers" of ggplot2
## 2) Graph exercises with our own ideas
```


```{r, echo=FALSE}
# Load standard packages
library(tidyverse)
```

## Introduction
We will look at the basic ggplot2 use using the faithful dataset, giving 
information on the eruption pattern of the Old Faithful geyser in Yellowstone
National Park.

## What is the grammar of graphics?

Graphics are all made of 8 constituents:
1. Theme: everything that is visual but that is not related to data (background colour, title, legend's title, etc.)

2. Coordinates: defines the physical mapping of the aesthetics to the graph (everything related to position)

3. Facets: Possibility of spling our plot area into different groups

4. Geometries: how to interpret aesthetics as graphical representations - how you specify (lines, points, box-plot)

5. Mapping: link data to graphical properties in the geometry (translates: this column is X, this column is colour)

6. Data: grammar of graphics requires a tidy format (the core of ggplot2 and any graphic technique)

```{r}
data("faithful")

faithful <- faithful %>% 
  as_tibble()

library(ggplot2)

# Which dataset to plot (is the "data =" argument)
ggplot(data = faithful, 
       
       ## Which columns to use for x and y
       mapping = aes(x = eruptions, y = waiting)) + 
  
       ## How do we want R to draw our plot
  geom_point()


ggplot(data = faithful,
       mapping = aes(x = eruptions, y = waiting)) +
  geom_point()

## The plus ("+") sign is used to combine ggplot2 elements (is the ggplot2 version of the pipe operator)

# Data and mapping can be given both as global (in ggplot()) or per layer
ggplot(faithful) + 
  geom_point(mapping = aes(x = eruptions, y = waiting))


```

If an aesthetic is linked to data it is put into `aes()`

```{r}
ggplot(faithful) + 
  # Mapping colour
  geom_point(aes(x = eruptions, y = waiting, colour = eruptions))

ggplot(faithful) + 
  # Mapping colour with a condition (R will understand that you want him to graph no the column anymore, but the condition)
  geom_point(aes(x = eruptions, y = waiting, colour = eruptions < 3))
```

If you simple want to set it to a value, put it outside of `aes()`

```{r}
## There is a different thing we can do with colour: instead of mapping the colour to a column, we want to set the colour
ggplot(faithful) + 
  geom_point(aes(x = eruptions, y = waiting),
## It needs to be outside our aesthetics (the legend its gone = no association between cols and colors anymore)
             colour = 'steelblue')
```

Some geoms only need a single mapping and will calculate the rest for you

```{r}
ggplot(faithful) + 
  geom_histogram(aes(x = eruptions))

## Geoms are drawn in the order they are added. The point layer is thus drawn on
## top of the histogram in the example below.

ggplot(faithful) + 
  geom_histogram(aes(x = eruptions)) +
  geom_point(aes(x = eruptions, y = waiting))

```

#### Exercise

Modify the code below to make the points larger squares and slightly transparent.
See `?geom_point` for more information on the point layer.

```{r}
## Shape as square: square = 0
ggplot(faithful) + 
  geom_point(aes(x = eruptions,
                 y = waiting),
             alpha = 0.85,
             shape = 19)

```

Hint 1: transparency is controlled with `alpha`, and shape with `shape`
Hint 2: remember the difference between mapping and setting aesthetics

* * *

Colour the two distributions in the histogram with different colours

```{r}
ggplot(faithful) + 
  geom_histogram(aes(x = eruptions, fill = eruptions < 3.1))
```

Hint 1: For polygons you can map two different colour-like aesthetics: `colour` 
(the colour of the stroke) and `fill` (the fill colour)

Fill is a different way of ggplot2 coloring (colour of the interior).

* * *

Colour the distributions in the histogram by whether `waiting` is above or below 
`60`. What happens?

```{r}
ggplot(faithful) + 
  geom_histogram(aes(x = eruptions))
```

Change the plot above by setting `position = 'dodge'` in `geom_histogram()` 
(while keeping the colouring by `waiting`). What do `position` control? 

* * *

Add a line that separates the two point distributions. See `?geom_abline` for 
how to draw straight lines from a slope and intercept.

```{r}
ggplot(faithful) + 
  geom_point(aes(x = eruptions, y = waiting))
```

### Scales
Scales define how the mapping you specify inside `aes()` should happen. All 
mappings have an associated scale even if not specified.

Preparing the data

```{r}
library(gapminder)
gapminder2 <- gapminder %>% 
  group_by(continent, year) %>% 
  summarize(avg_lifeExp = mean(lifeExp),
            avg_pop = mean(pop))
```


```{r}
ggplot(gapminder2) + 
  geom_point(aes(x = year, y = avg_lifeExp, colour = continent))
```

Positional mappings (x and y) also have associated scales.

```{r}
ggplot(gapminder2) + 
  geom_point(aes(x = year, y = avg_lifeExp, colour = continent)) +
  scale_y_continuous(breaks = c(20,30,40,50,60,70,80, 120), limits = c(20,120))
```

### Facets

The facet defines how data is split among panels. The default facet 
(`facet_null()`) puts all the data in a single panel, while `facet_wrap()` allows you to specify different types of small multiples grids.

```{r}
ggplot(gapminder2) + 
  geom_point(aes(x = year, y = avg_lifeExp, colour = continent)) +
  facet_wrap(~continent)

```

### Theme
Theming defines the feel and look of your final visualisation and is something
you will normally defer to the final polishing of the plot. It is very easy to 
change looks with a prebuild theme

```{r}
ggplot(gapminder2) + 
  geom_point(aes(x = year, y = avg_lifeExp, colour = continent)) +
  theme_minimal() + 
  labs(title = "Gapminder dataset",
       subtitle = "Subtitle",
       y = "Y-axis title",
       fill = "New Legend Name",
       x = "X-axis title",
       caption = "Caption") +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 18))
```

Exercise

Let's try something with ggplot2 x gapminder dataset!

```{r}
## gdpPercap vs lifeExp seems useful
gapminder %>% 
  filter(year %in% c(1952, 2007) & gdpPercap < 100000) %>% 
  ggplot() + 
  geom_point(aes(x = lifeExp, y = gdpPercap, colour = continent, size = pop)) +
  facet_wrap(~year)

  

```

